### 2020-07-10 implementation.tex
fix
related work

Hazel uses an optimizing pretty printer to determine layout, an approach 
based on the work of Bernardy et al \todo{cite bernardy}. This system relies 
fundamentally on character counts. Consequently, our implementation asks each 
livelit to specify dimensions in terms of character counts rather than pixels.
Livelits can be laid out in one of two ways: as inline livelits, like \li{\$slider},
which must be exactly one character high and appear inline with the code,
or as \emph{multi-line livelits}, which occupy up to the full width of the editor 
and a specified number of code lines. The pretty printer lays out multi-line livelits
on their own line, as seen in the figures in this paper.

We are currently exploring a more flexible system whereby 
multi-line livelits can be named and applied within a larger expression, but
the GUI is not displayed until the next line, even if there are intervening characters.
This is inspired by the \li{~} placeholder for large textual literals in the 
Wyvern programming language \cite{TSLs}. We are also considering a number of other layout 
options, e.g. pop-up livelits, livelits pinned to sidebars, and livelits that are rendered  
on a separate canvas or document while still formally being located within an underlying functional program.

This latter option 
would be particularly interesting for end-user programming scenarios: users with limited
programming experience 
could interact with a collection of livelits laid out separately, without necessarily
even being aware that their interactions are actually edits to an underlying typed
functional program. More experienced users on the team could then make programmatic use of the entered
data and parameters in a seamless manner. We are also excited about the possibility of end users
getting curious enough to ``View Source'' in this setting, and in so doing 
being exposed to typed functional programming concepts gradually.

The layout of splices within livelits is also an interesting problem, and our current approach, where 
the provider must specify a fixed size for each splice, is not particularly flexible. 
A more principled system for GUI layout that integrates with program pretty 
printing could be of substantial use in some settings. Formula bars, like the one in the \li{\$dataframe}
livelit, are a useful design pattern to avoid splice layout issues, and it may be possible to provide 
a generalized formula bar in the editor user interface that multiple livelits could share.
While splices can in principle contain other livelits, it is difficult for small splice editors to 
accomodate large livelit views. Fortunately, variables can be used to support composition via substitution 
rather than nesting. A linter might suggest such a refactoring.

@@

Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
especially with the edit awareness that Hazel provides. We leave further optimizations along these lines to future work.

>>>>

Hazel uses an optimizing pretty printer to determine layout, an approach 
based on the work of Bernardy et al \todo{cite bernardy}. This system relies 
fundamentally on character counts. Consequently, our implementation asks each 
livelit to specify dimensions in terms of character counts rather than pixels.
Livelits can be laid out in one of two ways: as inline livelits, like \li{\$slider},
which must be exactly one character high and appear inline with the code,
or as \emph{multi-line livelits}, which occupy up to the full width of the editor 
and a specified number of code lines (not shown in Fig.~\ref{fig:color-impl}). 
The pretty printer lays out multi-line livelits
on their own line, as seen in the figures in this paper.

We are currently exploring a more flexible system whereby 
multi-line livelits can be named and applied within a larger expression, but
the GUI is not displayed until the next line, even if there are intervening characters.
This is inspired by the \verb|~| placeholder for large textual literals in the 
Wyvern programming language \cite{TSLs}. We are also considering a number of other layout 
options, e.g. pop-up livelits, livelits pinned to sidebars, and livelits that are rendered  
on a separate canvas or document while still formally being located within an underlying functional program.

This latter option 
would be particularly interesting for end-user programming scenarios: users with limited
programming experience 
could interact with a collection of livelits laid out separately in the popular ``dashboard'' style, 
without necessarily
even being aware that their interactions are actually edits to an underlying typed
functional program. More experienced users on the team could then make programmatic use of the entered
data and parameters in a seamless manner. We are also excited about the possibility of end users
getting curious enough to ``View Source'' in this setting, and in so doing 
being exposed to typed functional programming concepts gradually.

The layout of splices within livelits is also an interesting problem, and our current approach, where 
the provider must specify a fixed size for each splice, is not particularly flexible. 
A more principled system for GUI layout that integrates with program pretty 
printing could be of substantial use in some settings. Formula bars, like the one in the \li{\$dataframe}
livelit, are a useful design pattern to avoid splice layout issues, and it may be possible to provide 
a generalized formula bar in the editor user interface that multiple livelits could share.
While splices can in principle contain other livelits, it is difficult for small splice editors to 
visually accomodate large livelit views in practice. 
Fortunately, variables can be used to support composition via substitution 
rather than nesting. A linter might suggest a refactoring to a variable in these scenarios.

@@

Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
especially with the edit awareness that Hazel provides. 
We leave further optimizations along these lines to future work.
Prior work on incrementally updating UIs as code changes could be relevant to this effort \cite{burckhardt2013s}.
